---
title: Soil spectral predictions of cereal grain ionomes from Malawi
author: A.M. Sila, M.G. Walsh, D.B. Kumssa, A.R. Mossa, T. Amede, P.C. Nalivata, D. Gashu, S.M. Haefele, S. Gameda, M.R. Broadley, E.J.M. Joy, E.H. Bailey, R.M. Lark, B.A. Walsh, R. Rodríguez Iglesias, J. Chamberlin and S. Snapp
date: "`r format(Sys.time(), '%d, %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_md: true
    number_sections: true
    css: style1.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Introduction

Micronutrients i.e., the essential vitamins and minerals, are important for child development as well as for the prevention of morbidity and disease in adults. Except for vitamin D, micronutrients are not produced in the body and must be obtained from the diet. Although people only need small amounts of micronutrients, consuming the [recommended daily amounts](https://www.fda.gov/food/new-nutrition-facts-label/daily-value-new-nutrition-and-supplement-facts-labels) is important as micronutrient deficiencies (MNDs) can have serious consequences on both health and well-being. "*Effective MND surveillance is essential to inform policy responses to current food system challenges and socioeconomic and environmental drivers of change. Potential food system interventions to alleviate MNDs, such as food fortification and biofortification through breeding crops with increased micronutrient concentrations in their edible portions, should consider subnational geographical effects, which can be larger in magnitude than planned intervention outcomes*" ([Gashu et al., 2020](https://www.nature.com/articles/s41586-021-03559-3)).

However, national surveys containing staple crop micronutrient data are not routinely collected in most developing countries because they are expensive, time-consuming, and limited to the annual harvest season(s), both in the field and the laboratory. Also, the databases that are needed to analyze dietary micronutrient supplies against demands should be georeferenced to be effective, but coordinates and time stamps are often either only roughly linked to sampled administrative or project regions or are completely absent in Africa. Similarly, agronomic interventions that focus on plant micronutrients for improving cropland and livestock productivity are inadequately supported by spatial and temporal data because the distributions of different staple crops and livestock are usually not known and are not regularly monitored at national scales.

Spectral signatures of soils, plants, animal tissues, and materials generally are defined by their reflectance or absorbance as a function of wavelength in the electromagnetic spectrum. Under laboratory conditions, the signatures result from electronic transitions of atoms and vibrational stretching and bending of structural groups of atoms that form molecules or crystals. Spectroscopy has been shown to provide highly repeatable, rapid, and low-cost measurements of different soil, plant, and animal tissue properties in numerous studies. The amount of light absorbed by a rock, soil, plant or animal tissue sample can be measured with minimal sample preparation (primarily drying and grinding) across a wide range of ultra-violet (UV), visible (VIS), near (NIR), and mid-infrared (MIR) wavelengths to provide a unique spectral signature. An individual measurement can be performed in about 20 seconds, in stark contrast to more conventional soil and plant tests, which are typically slow, labor-intensive, expensive, and/or use hazardous chemicals, which have to be handled carefully by laboratories.

This notebook examines the possibility for using MIR-based spectral predictions of mineral nutrient deficiency risks for humans using cereal grain data from Malawi. It is intended to be a companion to a similar spatial prediction workflow; the most recent version of which can be downloaded [here](https://osf.io/7xzqv). The data we will be using were generated by the [GeoNutrition Project](http://www.geonutrition.com). Nationally representative cropland sampling frames and the laboratory methods that were used in the data collections are well described in [Gashu et al., (2021)](https://doi.org/10.1038/s41586-021-03559-3). An overall data descriptor article provides additional information and some of the raw data downloads at: [Kumssa et al., (2022) ](https://www.nature.com/articles/s41597-022-01500-5). This notebook is maintained on GitHub ([here](https://github.com/mgwalsh/GeoNutrition/blob/main/Soil_spec_preds_crop_ionomes.Rmd)). Feel free to alter and use it as you see fit, but also please cite it when you do so. It is copyrighted as [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/).

# Data setup

To actually run this notebook, you will need to load the packages indicated in the chunk below.

```{r, results = 'hide'}
# Package names
packages <- c("tidyverse", "rgdal", "leaflet", "htmlwidgets", "DT", "compositions", "doParallel",
              "caret", "caretEnsemble", "quantreg")

# Install packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))
```

## Data downloads

The primary ionomic labels used in this notebook include *Inductively Coupled Plasma Mass Spectrometry* ([ICP-MS](https://en.wikipedia.org/wiki/Inductively_coupled_plasma_mass_spectrometry)) nutrient concentration measurements of cereal grain samples (in mg kg^-1^), including: Na, Mg, P, S, K, Ca, Cr, Mn, Fe, Co, Cu, Zn, Se, Mo, and I. They also include the georeference of where the samples were collected (lon/lat, EPSG:3857), the cereal crop that was sampled at that location, farmer reported fertilizer use (no, unknown, yes) and the provenance of the grain samples (i.e., from the standing crop, a field stack or a closely located grain store). Note that any crop nutrient measurements below their respective limits of detection were set to 2/3 of the respective minimum observed values. You can download the file from our OSF repository using the following link: [`MWI_grain_CoDa.csv`](https://osf.io/zj5b7).

The associated soil samples were air-dried and then finely ground to powder (<100 microns) using an agate sample mill. The samples were then loaded into aluminum microtiter plates (Bruker Optics, A752-96) using a micro spatula to fill the 6-mm diameter wells and leveled. Samples were loaded into four replicate wells, each scanned 32 times in MIR reflectance mode using a Bruker Tensor 27 MIR spectrometer with a high throughput screening arm and a liquid nitrogen cooled HgCdTe detector. The absorbance spectra obtained for each sample were then averaged and combined into a single flat dataframe. You can download the spectral feature file from our OSF repository via this link [`MWI_soil_mir_spectra.csv`](https://osf.io/w5ng6). Ensure that you place both those files into your designated working directory. The next chunk loads the ICP-MS data, and flags the respective limits of detection for reference.

```{r}
# Read the ICP-MS crop data
cpdat <- read.table("MWI_grain_CoDa.csv", header=T, sep=",")

# Detection limit tags (for reference)
cpdat$dNa <- ifelse(cpdat$Na > min(cpdat$Na), 1, 0)
cpdat$dMg <- ifelse(cpdat$Mg > min(cpdat$Mg), 1, 0)
cpdat$dP <- ifelse(cpdat$P > min(cpdat$P), 1, 0)
cpdat$dS <- ifelse(cpdat$S > min(cpdat$S), 1, 0)
cpdat$dK <- ifelse(cpdat$K > min(cpdat$K), 1, 0)
cpdat$dCa <- ifelse(cpdat$Ca > min(cpdat$Ca), 1, 0)
cpdat$dCr <- ifelse(cpdat$Cr > min(cpdat$Cr), 1, 0)
cpdat$dMn <- ifelse(cpdat$Mn > min(cpdat$Mn), 1, 0)
cpdat$dFe <- ifelse(cpdat$Fe > min(cpdat$Fe), 1, 0)
cpdat$dCo <- ifelse(cpdat$Co > min(cpdat$Co), 1, 0)
cpdat$dCu <- ifelse(cpdat$Cu > min(cpdat$Cu), 1, 0)
cpdat$dZn <- ifelse(cpdat$Zn > min(cpdat$Zn), 1, 0)
cpdat$dSe <- ifelse(cpdat$Se > min(cpdat$Se), 1, 0)
cpdat$dMo <- ifelse(cpdat$Mo > min(cpdat$Mo), 1, 0)
cpdat$dI <- ifelse(cpdat$I > min(cpdat$I), 1, 0)

# Count of the number of detectable nutrients ... out of 15
cpdat$dlim <- rowSums(cpdat[, 23:37])
```

```{r, echo = F, fig.align = "center", fig.cap = "**Fig. 1**: Cumulative distribution of the number of detectable nutrient elements per grain sample."}

par(pty="s", mar=c(4,4,1,1))
plot(ecdf(cpdat$dlim), main="", xlab="No. of detectable elements", ylab="ECDF", xlim=c(8, 15),
     verticals=T, lty=1, lwd=2, cex.lab=1.3, col="tomato", do.points=F)
abline(0.5,0, lty=1, col="grey")
```

## Initial compositional model

We run a compositional main effects regression (see: [van den Boogaart and Tolosana-Delgado, 2013](https://link.springer.com/book/10.1007/978-3-642-36809-7)) using maize as the designated control crop, and the detectable number of elements (`dlim`, see Fig. 1) in the recorded ICP-MS grain measurements. This quantifies the average ionomic differences between crop species conditioned on fertilizer use (no, unknown, yes) and the number of detectable elements, which can be used as a baseline for the spectral models that are presented in the subsequent sections. Section 3.5 augments this initially simple model with spectral predictions.

```{r}
y = acomp(cpdat[,8:22])
crop <- cpdat$crop
fert <- cpdat$fert
dlim <- cpdat$dlim

# Fit compositional base model
(base.lm <- lm(clr(y) ~ crop + fert + dlim))
```

The next chunk transforms the model coefficients back to the original compositional scale. 

```{r}
(coefs <- clrInv(coef(base.lm), orig=y))
```

## Spectral feature pre-processing

To speed up processing of spectral features for machine learning, it is common to initially to reduce the dimensionality and collinearity of the (spectral) data with methods including: [Principal components analysis (PCA)](https://en.wikipedia.org/wiki/Principal_component_analysis), [Independent components analysis (ICA)](https://en.wikipedia.org/wiki/Independent_component_analysis) as well as other signal processing techniques such as [Non-negative matrix factorization (NMF)](https://en.wikipedia.org/wiki/Non-negative_matrix_factorization) and [wavelet compression](https://en.wikipedia.org/wiki/Wavelet_transform), among others. We use PCA here; however, feel to experiment. The next chunk appends the resulting principal component scores to the grain ICP-MS data frame.

```{r, results = 'hide'}
# Load the spectral data
spdat <- read.table("MWI_soil_mir_spectra.csv", header=T, sep=",")
spdat <- na.omit(spdat) ## includes only complete cases

# Soil spectral principal components
spdat.pca <- prcomp(spdat[ ,2:3577], center=T, scale=T) ## centered and scaled soil spectral PCAs
spca <- predict(spdat.pca, spdat)
spca <- spca[ ,1:10] ## save the first 10 components, which explain >95% of the total soil MIR variability
ssid <- spdat["ssid"]
spca <- as.data.frame(cbind(ssid, spca))

# Save PCA model for reuse
fname <- paste("./MW_coda/", "MW", "_spec_pca.rds", sep = "")
saveRDS(spdat.pca, fname) ## saves soil spectral PCA model

# Merge crop and soil spectral (pca) files
cpdat <- merge(cpdat, spca, by="ssid")
```

# Spectral predictions

## Training / testing setup

We initially set a 90/10% training and testing partition. The function `createDataPartition` in the [`caret`](https://topepo.github.io/caret/index.html) package can be used to generate balanced splits of the data. We partition on the number of detectable elements variable (`dlim`) here.

```{r}
# Calculate the centered log ratios
varc <- c("Na", "Mg", "P", "S", "K", "Ca", "Cr", "Mn", "Fe", "Co", "Cu", "Zn", "Se", "Mo", "I")
comps <- cpdat[varc]
comps <- comps / rowSums(comps) ## close the composition
comps <- as.data.frame(clr(comps)) ## centered log ratio (clr) transform

# Attach survey and spectral data
locv <- c("ssid", "crop", "dlim", "PC1", "PC2", "PC3", "PC4", "PC5", "PC6", "PC7", "PC8", "PC9", "PC10") 
locs <- cpdat[locv]
coda <- cbind(locs, comps)
coda <- na.omit(coda) ## includes only complete cases

# Set calibration/validation set randomization seed
seed <- 12358
set.seed(seed)
coda <- coda[sample(1:nrow(coda)), ] # randomize observations

# Split data into training and test sets
gnIndex <- createDataPartition(coda$dlim, p = 9/10, list = F, times = 1)
gn_cal <- coda[ gnIndex,]
gn_val <- coda[-gnIndex,]
```

## Base-learner training

This chunk fits 3 base-learner models (one bagging, one boosting, one rule-based), which uses the spectral training data with repeated 10-fold cross-validation. You can learn more about how these algorithms work by looking at the following links: [`randomForest`](https://www.rdocumentation.org/packages/randomForest/versions/4.6-14/topics/randomForest), [`gbm`](https://www.rdocumentation.org/packages/gbm/versions/2.1.8.1) and [`cubist`](https://www.rdocumentation.org/packages/Cubist/versions/0.4.1). Change these as you see fit. There are about 160 different algorithms available via the [`caret`](https://www.rdocumentation.org/packages/caret/versions/6.0-93) and [`caretEnsemble`](https://www.rdocumentation.org/packages/caretEnsemble/versions/2.0.1) packages. 

```{r, warning = FALSE, results='hide'}
# Calibration labels
labs <- c("P") ## **insert other labels here** 
lcal <- as.vector(t(gn_cal[labs]))

# Spectral calibration features
fcal <- gn_cal[ ,c(3:13)]

# Start doParallel to parallelize model fitting
set.seed(seed)
mc <- makeCluster(detectCores())
registerDoParallel(mc)

# Specify model training controls
tc <- trainControl(method="repeatedcv", number=10, repeats=3, allowParallel = TRUE, savePredictions="final")

# Fit 3 base-learners using the specified ionomic labels
blist <- caretList(fcal, lcal,
                   trControl = tc,
                   tuneList = NULL,
                   methodList = c("rf", "gbm", "cubist"),
                   preProcess = c("center","scale"),
                   metric = "RMSE")

stopCluster(mc)
fname <- paste("./MW_coda/", labs, "_blist.rds", sep = "")
saveRDS(blist, fname)

# Generate predictions (**change the labels manually here**)
gn_cal$P_rf <- predict(blist$rf, gn_cal)
gn_cal$P_gb <- predict(blist$gbm, gn_cal)
gn_cal$P_cu <- predict(blist$cubist, gn_cal)

# Same for the test-set
gn_val$P_rf <- predict(blist$rf, gn_val)
gn_val$P_gb <- predict(blist$gbm, gn_val)
gn_val$P_cu <- predict(blist$cubist, gn_val)
```

```{r, echo = FALSE, results = 'hide'}
# write.csv(gn_cal, "./MW_coda/MW_base_spec_cal.csv", rownames = FALSE)
# write.csv(gn_val, "./MW_coda/MW_base_spec_val.csv", rownames = FALSE)
```

To save time, we only demonstrate the fitting processses here (on phosphorus), but you can run the code on any of the other nutrients by switching labels in the places indicated in the script. We have pre-trained all 45 base plus the subsequent 15 stacked learners (section 3.3) that we shall be using for the model testing and prediction rule generation steps below. This is a bit of hack! However, you can download and check all of the associated `.Rds` model objects and dataframes from our OSF repository [here](https://osf.io/49px5/). To run the next sections you have to download the needed files using the following links: [`MW_stack_spec_cal.csv`](https://osf.io/gw6uv) and [`MW_stack_spec_val.csv`](https://osf.io/2qem8). Again, please make sure to place those into your designated working directory.

```{r}
# Scrub extraneous objects in memory
rm(list=setdiff(ls(), c("cpdat")))

# Read the base-learner training and testing files
gn_cal <- read.table("MW_stack_spec_cal.csv", header=T, sep=",")
gn_cal[ , c('X.1', 'X')] <- list(NULL)
gn_val <- read.table("MW_stack_spec_val.csv", header=T, sep=",")
gn_val[ , c('X.1', 'X')] <- list(NULL)
```

## Stacked-learner training

This next chunk then fits and predicts ensemble regressions based on the 45 base-learner models, using the training data.

```{r, results = 'hide'}
# Calibration labels
labs <- c("P") ## **insert other labels here** 
lcal <- as.vector(t(gn_cal[labs]))

# Base-learner calibration features
fcal <- gn_cal[ ,c(30:74)]

# Start doParallel to parallelize model fitting
set.seed(1235813)
mc <- makeCluster(detectCores())
registerDoParallel(mc)

# Specify model training controls
tc <- trainControl(method="repeatedcv", number=10, repeats=3, allowParallel = TRUE, savePredictions="final")

# Model training
en <- train(fcal, lcal,
            method = "glmStepAIC",
            trControl = tc,
            metric = "RMSE")

stopCluster(mc)
fname <- paste("./MW_coda/", labs, "_ens.rds", sep = "")
saveRDS(en, fname)

# Generate predictions (**change the labels manually here**)
gn_cal$P_en <- predict(en, gn_cal)

# Same for the test-set
gn_val$P_en <- predict(en, gn_val)
```

```{r, echo = FALSE, results = 'hide'}
# write.csv(gn_cal, "./MW_coda/MW_stack_spec_cal.csv", rownames = FALSE)
# write.csv(gn_val, "./MW_coda/MW_stack_spec_val.csv", rownames = FALSE)
```

## Prediction uncertainties

The models that have been developed have not seen any of the test-set (evaluation) data up to this point. While there are many ways to quantify the uncertainty inherent in the ensemble predictions, we take a simple but quite robust approach here using quantile regression with ([quantreg](https://cran.r-project.org/web/packages/quantreg/quantreg.pdf)). We are mainly interested in the initial spread of the [ROI-wide](https://en.wikipedia.org/wiki/Region_of_interest) predictions (sensu, their 80% probable intervals).

```{r, results = 'hide'}
# Quantile regression test-set fits
qNa <- rq(Na ~ Na_en, tau=c(0.1,0.5,0.9), data = gn_val)
qMg <- rq(Mg ~ Mg_en, tau=c(0.1,0.5,0.9), data = gn_val)
qP <- rq(P ~ P_en, tau=c(0.1,0.5,0.9), data = gn_val)
qS <- rq(S ~ S_en, tau=c(0.1,0.5,0.9), data = gn_val)
qK <- rq(K ~ K_en, tau=c(0.1,0.5,0.9), data = gn_val)
qCa <- rq(Ca ~ Ca_en, tau=c(0.1,0.5,0.9), data = gn_val)
qCr <- rq(Cr ~ Cr_en, tau=c(0.1,0.5,0.9), data = gn_val)
qMn <- rq(Mn ~ Mn_en, tau=c(0.1,0.5,0.9), data = gn_val)
qFe <- rq(Fe ~ Fe_en, tau=c(0.1,0.5,0.9), data = gn_val)
qCo <- rq(Co ~ Co_en, tau=c(0.1,0.5,0.9), data = gn_val)
qCu <- rq(Cu ~ Cu_en, tau=c(0.1,0.5,0.9), data = gn_val)
qZn <- rq(Zn ~ Zn_en, tau=c(0.1,0.5,0.9), data = gn_val)
qSe <- rq(Se ~ Se_en, tau=c(0.1,0.5,0.9), data = gn_val)
qMo <- rq(Mo ~ Mo_en, tau=c(0.1,0.5,0.9), data = gn_val)
qI <- rq(I ~ I_en, tau=c(0.1,0.5,0.9), data = gn_val)
```

```{r, echo = FALSE, fig.align = "center", fig.show = 'hold', fig.height = 12, fig.cap = "**Fig. 2**: Quantile regression plots of predicted vs measured test-set values (on a centered log ratio scale). The red lines are estimated at the median, the blue lines are estimated at the 10% and 90% quantiles."}

colors <- c("dodger blue", "tomato", "dodger blue")
par(mfrow = c(5,3), pty = "s", mar=c(2,4,1,1))

plot(Na ~ Na_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qNa$coefficients)) {
    abline(coef(qNa)[, j], col = colors[j], lwd = 2)}

plot(Mg ~ Mg_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qMg$coefficients)) {
    abline(coef(qMg)[, j], col = colors[j], lwd = 2)}

plot(P ~ P_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qP$coefficients)) {
    abline(coef(qP)[, j], col = colors[j], lwd = 2)}

plot(S ~ S_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qP$coefficients)) {
    abline(coef(qP)[, j], col = colors[j], lwd = 2)}

plot(K ~ K_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qP$coefficients)) {
    abline(coef(qP)[, j], col = colors[j], lwd = 2)}

plot(Ca ~ Ca_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qP$coefficients)) {
    abline(coef(qP)[, j], col = colors[j], lwd = 2)}

plot(Cr ~ Cr_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qP$coefficients)) {
    abline(coef(qP)[, j], col = colors[j], lwd = 2)}

plot(Mn ~ Mn_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qP$coefficients)) {
    abline(coef(qP)[, j], col = colors[j], lwd = 2)}

plot(Fe ~ Fe_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qP$coefficients)) {
    abline(coef(qP)[, j], col = colors[j], lwd = 2)}

plot(Co ~ Co_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qP$coefficients)) {
    abline(coef(qP)[, j], col = colors[j], lwd = 2)}

plot(Cu ~ Cu_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qP$coefficients)) {
    abline(coef(qP)[, j], col = colors[j], lwd = 2)}

plot(Zn ~ Zn_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qP$coefficients)) {
    abline(coef(qP)[, j], col = colors[j], lwd = 2)}

plot(Se ~ Se_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qP$coefficients)) {
    abline(coef(qP)[, j], col = colors[j], lwd = 2)}

plot(Mo ~ Mo_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qP$coefficients)) {
    abline(coef(qP)[, j], col = colors[j], lwd = 2)}

plot(I ~ I_en, data = gn_val, xlab="", col = "dark grey", cex.lab = 1.3)
for (j in 1:ncol(qP$coefficients)) {
    abline(coef(qP)[, j], col = colors[j], lwd = 2)}
```

## Ionomic prediction rules

This final chunk is intended to generate a foundation (at least from the computing side) for the development of a context-based [recommender system](https://en.wikipedia.org/wiki/Recommender_system) that would allow users to make evidence based decisions about crop selections, biofortification and/or mineral nutrient supplementation practices in different soil environments. This is a spectrally augmented version of the model that was presented in section 2.2.  

```{r}
# Compositional setup
coda <- rbind(gn_cal, gn_val)
coda <- coda[, c(1, 74:88)]
cpdat <- merge(cpdat, coda, by = 'ssid')
y = acomp(cpdat[,8:22])

# Select covariates
crop <- cpdat$crop
Na_en <- cpdat$Na_en
Mg_en <- cpdat$Mg_en
P_en <- cpdat$P_en
S_en <- cpdat$S_en
K_en <- cpdat$K_en
Ca_en <- cpdat$Ca_en
Cr_en <- cpdat$Cr_en
Mn_en <- cpdat$Mn_en
Fe_en <- cpdat$Fe_en
Co_en <- cpdat$Co_en
Cu_en <- cpdat$Cu_en
Zn_en <- cpdat$Zn_en
Se_en <- cpdat$Se_en
Mo_en <- cpdat$Mo_en
I_en <- cpdat$I_en

# Fit compositional regression
spec_rules.lm <- lm(clr(y) ~ crop + Na_en + Mg_en + P_en + S_en + K_en + Ca_en + Cr_en +
                    Mn_en + Fe_en + Co_en + Cu_en + Zn_en + Se_en + Mo_en + I_en)
summary(spec_rules.lm)
```

# Main takeaways

* This notebook is intended to be a companion to a similar notebook, which covers spatial prediction workflows ([download here](https://osf.io/7xzqv)). In fact, the two approaches could be deployed either in tandem or in sequence to guide and stratify new measurement campaigns, recommender validation, and evidence-based nutrient monitoring activities.

* The ensemble models that are presented here perform very well in cross-validation runs, and are not overfit ... in our opinion. However, because the current dataset is small (by machine learning standards), there are caveats with regard to how well the current predictions generalize to unseen or uncalibrated regions of interest (see Fig. 2). Our main intent for this notebook was to provide a computational example of how ionomic information might be gathered more effectively in the future. The spectral models would learn from new spectral data quickly and would be expected to improve and generalize to new regions of interest over time.

* Generating more field and laboratory data are not expected to be cost-prohibitive for either national governments nor for international donors and could be accomplished quickly. We will provide more information about this might be accomplished and will also suggest "*GeoNutrition*" standard operating procedures in a subsequent notebook.

* The "[low code](https://www.ibm.com/topics/low-code)" development approach that is taken here can be run on a normal (2.4 GHz, 8 core, 32 Gb memory) computer. This is expected to facilitate data and code reuse that promotes further innovation and operational applications development of [recommender systems](https://en.wikipedia.org/wiki/Recommender_system) in GeoNutrition projects in Africa.

Any questions, comments or corrections related to this notebook are most welcome via [AFSIS](mailto:mgw.africasoils.info), and we will attempt to include them in any updates.

